Supp mat: time series paper

### 2018-02-21 writing/manuscript_draft_short_version.tex
changes to short format version of the article

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \usepackage{lineno}
\usepackage{setspace}\doublespacing \usepackage[firstinits=true,style=authoryear-comp,natbib=true,doi=false,isbn=false,url=false,uniquename=false,uniquelist=false,sorting=none]{biblatex}

\usepackage{titlesec}

\titlespacing\section{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
\titlespacing\subsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}
\titlespacing\subsubsection{0pt}{-5pt plus 1pt minus 1pt}{-5pt plus 1pt minus 1pt}

@@

Running head: Power to detect population trends {[}33 of 40
characters{]}

@@

To be submitted to: \emph{Ecology} as a Statistical Report \vspace{1 mm}

Number of pages: 20\\
\vspace{1 mm}

@@

Data and code for all the figures and tables can be found at
(\url{https://github.com/erwhite1/time-series-project}). \vspace{2 mm}

@@

Observational studies and population time series have become a
cornerstone of modern ecological research and conservation biology
(Magurran et al. 2010; Hughes et al. 2017). Long-term data are necessary
to both understand population dynamics and to assess species extinction
risk. Some time series may now be considered ``long-term''
(e.g.~continuous plankton recorder, Giron-Nava et al. (2017)), but most
are still short. Time series are typically short due to short funding
cycles and typical experimental time-frames (Field et al. 2007; Hughes
et al. 2017).

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols and
Williams 2006). A short time series may lead to wrong conclusions given
large natural year-to-year variability (McCain, Szewczyk, and Knight
2016). Managers need to know when action is needed for a population.
Therefore, managers must understand when a population trend over time is
actually meaningful. For example, the International Union for
Conservation of Nature (IUCN) Red List Categories and Criteria suggest,
under Criterion A2, a species qualifies as vulnerable if it has
experienced a 30\% decline over 10 years, or three generations (IUCN
2012). Sampling is typically expensive, therefore, we also do not want
to sample for longer than is necessary. For example, Gerber, DeMaster,
and Kareiva (1999) investigated the minimum time series required to
estimate population growth of the endangered, but recovering, eastern
North Pacific gray whale (\emph{Eschrichtius robustus}). They used a
long-term census to retroactively determine the minimum time series
required to assess threat status. They found that only 11 years were
needed, eight years before the delisting decision was made. This
highlights the importance of estimating the minimum time series required
as an earlier decision would have saved time and money (Gerber,
DeMaster, and Kareiva 1999). Further, waiting too long to decide an
action can imperil a species where management action could have been
taken earlier (Martin et al. 2012; Martin et al. 2017).

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Wagner,
Vandergoot, and Tyson 2009; Giron-Nava et al. 2017). For example, Rhodes
and Jonzen (2011) examined the optimal allocation of effort between
spatial and temporal replicates. Using simple populations models, they
found that the allocation of effort depends on environmental variation,
spatial and temporal autocorrelation, and observer error. Rueda-Cediel
et al. (2015) also used a modeling approach, but parameterized a model
specific for a threatened snail, \emph{Tasmaphena lamproides}. They
found that for this short-lived organism, 15 years was adequate to
assess long-term trends in abundance. However, these studies, and other
past work, have typically focused either on theoretical aspects of
monitoring design or focused on only a few species.

I estimate the minimum time series length required (\(T_{min}\)) to
assess long-term changes in abundance via simple linear regression.
Statistical power is not a new tool (Cohen 1992), but it is still
under-appreciated in ecological research (Legg and Nagy 2006).
Therefore, I only briefly describe the idea of statistical power for
time series analyses using simulations. I then examine 822 population
time series to estimate \(T_{min}\).

\section{Methods}\label{methods}

\subsection{Statistical power in time
series}\label{statistical-power-in-time-series}

An important step in experimental design is to determine the number of
samples required. For any particular experiment four quantities are
intricately linked: significance level (\(\alpha\)), statistical power,
effect size, and sample size (Legg and Nagy 2006). The exact
relationship between these quantities depends on the specific
statistical test. A type I error is a false positive, or incorrect
rejection of a true null hypothesis. For example, if a time series was
assessed as significantly increasing or decreasing---when there was no
true significant trend---this would be a false positive. The false
positive rate, or significance level (\(\alpha\)) is often set at 0.05
(although this is purely historical, Mapstone (1995)). A type II error
(\(\beta\)) is a failure to detect a true trend, or failure to reject a
false null hypothesis. Formally, statistical power (\(1-\beta\)) is one
minus the probability of a type II error (\(\beta\)). The effect size is
a estimate of the strength of a particular phenomenon. Prior to an
experiment, one could set appropriate levels of power, significance
level, and the effect size to estimate the sample size required for the
experiment. This approach, however, is not straight-forward for a time
series, or more complicated scenarios (P. C. Johnson et al. 2015), as
data are clearly non-independent.

\subsection{Simulation approach}\label{simulation-approach}

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). Essentially, a population
model, with a selected set of parameter values, is simulated
repetitively for a number of years. As an example, we can take the
following population model for population size \(N\) at time \(t\):

@@

where \(\epsilon\) is a normally-distributed random noise term with mean
\(\mu\) and standard deviation \(\sigma\). The rate of growth (\(r\)) is
the trend strength of the increase or decrease (i.e.~the estimated slope
from linear regression). It is important to note that any population
model could be used here. As an example, in the supplementary material
(Fig. \ref{fig:shark_example}), an age-structured model of population
dynamics is used in place of equation \ref{sim_model} (White, Nagy, and
Gruber 2014).

Statistical power is the proportion of simulations where the slope
parameter from linear regression is significant at the 0.05 threshold.
Statistical power of 0.8 would indicate that, if there was a true trend
in abundance, there would be a 0.8 probability of detecting the trend.
Values of 0.05 for the significance level and 0.8 statistical power are
historical (Cohen 1992) and it is important to examine the effect of
changing these values (Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, the model in equation
\ref{sim_model} is simulated repetitively. Statistical power increases
with increases in the length of time sampled (Fig.
\ref{fig:theoretical_approach}b). Where power is greater than 0.8 (the
dotted line), that is the minimum time required (\(T_{min}\)) to be
confident in the detection of a long-term trend in abundance. As shown
previously (Rhodes and Jonzen 2011; Rueda-Cediel et al. 2015),
statistical power increases with larger trend strength and lower
population variability (Fig. \ref{fig:theoretical_approach}c,d).
Simulation approaches can be useful before designing a monitoring
program or when a realistic model exists for the population in question.
However, it is also informative to examine how statistical power changes
with sampling effort in empirical time series.

\subsection{Data source}\label{data-source}

I use a database of 2444 population time series compiled by Keith et al.
(2015). The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010) and several other
sources (Keith et al. 2015). I filtered out short time series (less than
35 years), and those with missing data, leaving 822 time series. The
data includes information on 477 vertebrate species with a focus on
mammals, birds, and fish.

\subsection{Empirical approach}\label{empirical-approach}

I assume that each time series is long enough to include all necessary
information (e.g.~variability) about the population. In other words,
each time series is a representative sample. I first take all possible
contiguous subsamples of each time series. For example, a time series of
35 years would have 34 possible contiguous subsamples of length 2, 34
possible contiguous subsamples of length 3, and continuing until 1
possible contiguous subsample of length 35 (Gerber, DeMaster, and
Kareiva 1999; Brashares and Sam 2005; Giron-Nava et al. 2017). In line
with the simulation approach, I determine the proportion of subsamples
of a particular length that have estimated slope coefficients
statistically different from zero. This proportion is a measure of
statistical power. Lastly, I determine which subsample length is
required to achieve a certain threshold of statistical power (0.8, Cohen
(1992)). The minimum subsampled length that met these criteria is the
minimum time series length required (\(T_{min}\)). All analyses were
conducted in R (R Core Team 2016).

\section{Results}\label{results}

I examined a database of 822 separate population time series
representing 477 species. This database consisted of vertebrate species
with a variety of life-history characteristics. Across all the
populations, I found an average minimum time series length required
(\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution (Fig.
\ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
Actinopterygii) typically had estimates of \(T_{min}\) over 20 years.
Birds (class Aves) had a much wider distribution of \(T_{min}\), but
usually required less years of sampling. Differences between these
classes can be explained by differences in variability in population
size and strength of trends in abundance (Fig. \ref{fig:class}).

@@

The minimum time series length required was strongly correlated with
trend strength (i.e.~estimated slope coefficient from linear
regression), coefficient of variation in population size, and
autocorrelation in population size (Figs. \ref{fig:correlates}a-c). All
three of these explanatory variables were significant and had large
effect sizes (see Table A1). Combined, trend strength, coefficient of
variation in population, and autocorrelation account for 75.1\% of the
explained deviance (Zuur et al. 2009) in minimum time series length
required.

There was life-history information available (Myhrvold et al. 2015) for
547 populations representing 315 different species, all of which were
birds (Aves class). Some life-history traits were significant predictors
for the minimum time required (Figs. \ref{fig:correlates}d-h, Tables
A2,A3). However, even combined, all five of the life-history traits
accounted for only 5.99\% of the explained deviance in minimum time
series length required. In addition, when accounting for trend strength,
coefficient of variation, and autocorrelation, no life-history traits
were significant predictors of the minimum time required (Table A3).

@@

\section{Discussion}\label{discussion}

I examined 822 population time series. I then subsampled each to
determine the minimum time required to achieve a desired significance
level and power for linear regression. Statistical power is important as
it provides information on the necessary number of samples required to
determine a significant trend (Legg and Nagy 2006). I found that on
average 15.91 years of continuous monitoring were typically necessary
(Fig. \ref{fig:min_time_dist}b). However, the distribution of minimum
time required was wide. Therefore, simplistic rules of thumb are
probably not appropriate. This time-frame is in line with past work on a
short-lived snail species (Rueda-Cediel et al. 2015) and a long-lived
whale species (Gerber, DeMaster, and Kareiva 1999). Hatch (2003) used
seabird monitoring data to estimate minimum sampling requirements. He
found that the time required ranged from 11 to 69 years depending on
species, trend strength, and study design.

In line with theoretical predictions (Rhodes and Jonzen 2011), I also
found \(T_{min}\) was strongly correlated with the trend strength,
variability in population size, and temporal autocorrelation (Fig.
\ref{fig:correlates}). Contrary to my prior expectations, I also found
that \(T_{min}\) did not correlate with any life-history traits (Fig.
\ref{fig:correlates}d-h). I initially hypothesized that species with
longer lifespans or generation times may require a longer sampling
period. This result could have been a result of at least two factors.
First, the data I used may not include a diverse enough set of species
with different life-history traits. Second, the question I posed,
whether a population is increasing or decreasing, was specifically
concerned with population trends over time. Therefore, life-history
characteristics may be more important for other questions more closely
tied to species biology. For example, Blanchard, Maxwell, and Jennings
(2007) used detailed simulations of spatially-distributed fisheries to
compare survey designs. They found that statistical power depended on
survey design, temperature preferences, and the degree of population
patchiness.

An important related idea is the optimal allocation of sampling effort
in space versus time. In a theoretical investigation of this question,
Rhodes and Jonzen (2011) found that the optimal allocation of sampling
depended strongly on temporal and spatial autocorrelation. If spatial
population dynamics were highly correlated, then it was better to sample
more temporally, and vice versa. My work supports this idea as
populations with strong temporal autocorrelation needed less years of
sampling (Fig. \ref{fig:correlates}). Morrison and Hik (2008) also
studied the optimal allocation of sampling effort in space versus time,
but used emprical data from a long-term survey of the collared pika
(\emph{Ochotona collaris}) in the Yukon. They found that surveys less
than 5 years may be misleading and that extrapolating from one
population to another, even when nearby geographically, may be
untenable.

Seavy and Reynolds (2007) asked whether statistical power was even a
useful framework for assessing long-term population trends. They used 24
years of census data on Red-tailed Tropicbirds
(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
decline over 10 years almost always resulted in high statistical power
(above 0.8). Therefore, they cautioned against only using power analyses
to design monitoring schemes and instead argued for metrics that would
increase precision: improving randomization, reducing bias, and
increasing detection probability. Power analyses should not be the only
consideration when designing monitoring schemes. However, unlike Seavy
and Reynolds (2007), the results here indicate that longer than 10 years
is often needed to achieve high statistical power.

As a test of the consequences of not conducting power analyses, I
compared the results here to the IUCN Redlist criteria. IUCN criteria A2
suggests that species that have experienced 30\% declines over 10 years
(or three generations) should be listed as vulnerable (IUCN 2012).
However, for the populations I examined, this criteria may be too
simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the
IUCN criteria suggest more years than necessary are required to assess a
population as vulnerable (points below diagonal line in Fig.
\ref{fig:IUCN_analysis}). Conversely, for other populations the IUCN
criteria suggest sampling times that are less than the minimum time
required for statistical power. This suggests that the IUCN criteria are
probably too simplistic as the minimum time required does not correlate
with generation time or other biological covariates (Fig.
\ref{fig:correlates}d-h).

The design of monitoring programs should include calculations of
statistical power, the allocation of sampling in space versus time
(Rhodes and Jonzen 2011), and metrics to increase precision (Seavy and
Reynolds 2007). Ideally, a formal decision analysis to evaluate these
different factors would be conducted to design or assess any monitoring
program (Hauser, Pople, and Possingham 2006; McDonald-Madden et al.
2010). This type of formal decision analysis would also include
information on the costs of monitoring. These costs include the actual
costs of sampling (Brashares and Sam 2005) and the ecological costs on
inaction (Thompson et al. 2000).

\subsection{Limitations}\label{limitations}

This paper has some limitations in determining the minimum time series
length required. First, \(T_{min}\) is particular to the specific
question of interest. An additional complication is that for the
empirical approach, the subsampling of the full time series allows for
estimates of power, but the individual subsamples are clearly not
independent of one another. In an ideal setting, a specific population
model would be parameterized for each population of interest. Then,
model simulations could be used to estimate the minimum time series
required to address each specific question of interest. Clearly, this is
not always practical, especially if conducting analyses for a wide array
of species.

\subsection{Conclusions}\label{conclusions}

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{float} \usepackage{lineno}
\usepackage{setspace}\doublespacing \usepackage[round]{natbib}
\bibpunct[; ]{(}{)}{,}{a}{}{,}

@@

@@

To be submitted to: \emph{BioScience} as a Forum \vspace{1 mm}

Number of words: 4,400 (21 pages)\\
\vspace{1 mm}

@@

All data and code is freely available at:
(\url{https://github.com/erwhite1/time-series-project}). \vspace{2 mm}

@@

Observational studies and population time series have become a
cornerstone of modern ecological research and conservation biology
(Magurran \emph{et al.} 2010, Peters 2010, Hughes \emph{et al.} 2017).
Long-term data are necessary to both understand population dynamics and
to assess species extinction risk. Some time series may now be
considered ``long-term'' (e.g.~continuous plankton recorder, Giron-Nava
\emph{et al.} (2017)), but most are still short. Time series are
typically short due to short funding cycles and typical experimental
time-frames (Field \emph{et al.} 2007, Hughes \emph{et al.} 2017).

How long of a time series is actually necessary? This question has
important implications for both research and management (Nichols \&
Williams 2006). A short time series may lead to wrong conclusions given
large natural year-to-year variability (McCain \emph{et al.} 2016).
Managers need to know when action is needed for a population. Therefore,
managers must understand when a population trend over time is actually
meaningful. For example, the International Union for Conservation of
Nature (IUCN) Red List Categories and Criteria suggest, under Criterion
A2, a species qualifies as vulnerable if it has experienced a 30\%
decline over 10 years, or three generations (IUCN 2012). Sampling is
typically expensive, therefore, we also do not want to sample for longer
than is necessary. For example, Gerber \emph{et al.} (1999) investigated
the minimum time series required to estimate population growth of the
endangered, but recovering, eastern North Pacific gray whale
(\emph{Eschrichtius robustus}). They used a long-term census to
retroactively determine the minimum time series required to assess
threat status. They found that only 11 years were needed, eight years
before the delisting decision was made. This highlights the importance
of estimating the minimum time series required as an earlier decision
would have saved time and money (Gerber \emph{et al.} 1999). Further,
waiting too long to decide an action can imperil a species where
management action could have been taken earlier (Martin \emph{et al.}
2012, 2017).

Past work has investigated questions related to the minimum time series
required to estimate trends in population size over time (Wagner
\emph{et al.} 2009, Giron-Nava \emph{et al.} 2017). For example, Rhodes
\& Jonzen (2011) examined the optimal allocation of effort between
spatial and temporal replicates. Using simple populations models, they
found that the allocation of effort depends on environmental variation,
spatial and temporal autocorrelation, and observer error. Rueda-Cediel
\emph{et al.} (2015) also used a modeling approach, but parameterized a
model specific for a threatened snail, \emph{Tasmaphena lamproides}.
They found that for this short-lived organism, 15 years was adequate to
assess long-term trends in abundance. However, these studies, and other
past work, have typically been only on theoretical aspects of monitoring
design or focused on a few species.

Statistical power is not a new tool (Cohen 1992, Gibbs \emph{et al.}
1998), but it is still under-appreciated in ecological research (Legg \&
Nagy 2006). Therefore, I begin by reviewing key concepts of power
analyses in relation to time series analysis. I then explain how
simulation approaches have been used to estimate the minimum time
required to estimate long-term population trends. Lastly, I take an
empirical approach to estimate the minimum time required for 822 animal
populations.

\section{Statistical power in time series
analyses}\label{statistical-power-in-time-series-analyses}

An important step in experimental design is to determine the number of
samples required. For any particular experiment four quantities are
intricately linked: significance level (\(\alpha\)), statistical power,
effect size, and sample size (Legg \& Nagy 2006). The exact relationship
between these quantities depends on the specific statistical test. A
type I error is a false positive, or incorrect rejection of a true null
hypothesis. For example, if a time series was assessed as significantly
increasing or decreasing---when there was no true significant
trend---this would be a false positive. The false positive rate, or
significance level (\(\alpha\)) is often set at 0.05 (although this is
purely historical, Mapstone (1995)). A type II error (\(\beta\)) is a
failure to detect a true trend, or failure to reject a false null
hypothesis. Formally, statistical power (\(1-\beta\)) is one minus the
probability of a type II error (\(\beta\)). The effect size is a
estimate of the strength of a particular phenomenon.

Prior to an experiment, one could set appropriate levels of power,
significance level, and the effect size to estimate the sample size
required for the experiment. This approach, however, is not
straight-forward for a time series, or more complicated scenarios
(Johnson \emph{et al.} 2015), as data are clearly non-independent.

In the context of time series data, sample size can be the number of
study sites surveyed, frequency of surveys per year, and the number of
years surveyed. For example, Gibbs \emph{et al.} (1998) examined how
many times within a year a population needs to be sampled to ensure high
statistical power. They found that the sampling intensity within a year
differed greatly depending on species, because of differences in
population variability. I explore similar methods, but I focus on the
number of years required to estimate trends in abundance.

\section{Simulation approach}\label{simulation-approach}

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987,
Gibbs \emph{et al.} 1998). This is the same approach one might use in
sample size calculations for any experimental design too complicated for
simple power analyses (Bolker 2008, Johnson \emph{et al.} 2015).
Essentially, a population model, with a selected set of parameter
values, is simulated repetitively for a number of years. As an example,
we can take the following population model for population size \(N\) at
time \(t\):

@@

where \(\epsilon\) is a normally-distributed random noise term with mean
\(\mu\) and standard deviation \(\sigma\). The rate of growth (\(r\)) is
the trend strength of the increase or decrease (i.e.~the estimated slope
from linear regression).

With regard to detecting time series trends, statistical power is the
proportion of simulations where the slope parameter from linear
regression is significant at the 0.05 threshold. Statistical power of
0.8 would indicate that, if there was a true trend in abundance, there
would be a 0.8 probability of detecting the trend. Values of 0.05 for
the significance level and 0.8 statistical power are historical (Cohen
1992) and it is important to examine the effect of changing these values
(Fig. \ref{fig:min_time_vs_alpha_beta}).

In Fig. \ref{fig:theoretical_approach}a, the model in equation
\ref{sim_model} was simulated repetitively. Statistical power increases
with increases in the length of time sampled (Fig.
\ref{fig:theoretical_approach}b). Where power is greater than 0.8 (the
dotted line), that is the minimum time required (\(T_{min}\)) to be
confident in the detection of a long-term trend in abundance. As shown
previously (Rhodes \& Jonzen 2011, Rueda-Cediel \emph{et al.} 2015),
statistical power increases with larger trend strength and lower
population variability (Fig. \ref{fig:theoretical_approach}c,d).
Simulation approaches can be useful before designing a monitoring
program or when a realistic model exists for the population in question.
However, it is also informative to examine how statistical power changes
with sampling effort in empirical time series.

It is important to note that any population model could be used here
(see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the
specific model choice should be tailored to the population of interest.
As an example, I determined the minimum time required to estimate
long-term population trends using a stochastic, age-structured model of
lemon shark population dynamics in the Bahamas (White \emph{et al.}
2014). I found that over 27 years of continuous monitoring were needed
in this particular scenario (Fig. \ref{fig:shark_example}). Not
surprisingly, the minimum time required for the lemon shark population
was strongly dependent on model parameters. Similarly, Rueda-Cediel
\emph{et al.} (2015) used a matrix model parameterized for a particular
snail species. They used the model to argue that only 10-15 years were
needed to accurately assess trends in abundance.

\section{Empirical approach}\label{empirical-approach}

As an empirical test of these ideas, I used a database of 2444
population time series compiled by Keith \emph{et al.} (2015). The data
are originally from the Global Population Dynamics Database (NERC Centre
for Population Biology 2010) and several other sources (Keith \emph{et
al.} 2015). I filtered out short time series (less than 35 years), and
those with missing data, leaving 822 time series. The data includes
information on 477 vertebrate species with a focus on mammals, birds,
and fish.

I assumed that each time series was long enough to include all necessary
information (e.g.~variability) about the population. In other words,
each time series was a representative sample. I first took all possible
contiguous subsamples of each time series. For example, a time series of
35 years had 34 possible contiguous subsamples of length 2, 34 possible
contiguous subsamples of length 3, and continuing until 1 possible
contiguous subsample of length 35 (Gerber \emph{et al.} 1999, Brashares
\& Sam 2005, Giron-Nava \emph{et al.} 2017). In line with the simulation
approach, I determined the proportion of subsamples of a particular
length that had estimated slope coefficients statistically different
from zero. This proportion is a measure of statistical power. Lastly, I
determined which subsample length is required to achieve a certain
threshold of statistical power (0.8, Cohen (1992)). The minimum
subsampled length that met these criteria was the minimum time series
length required (\(T_{min}\)). All analyses were conducted in R (R Core
Team 2017).

\subsection{Estimates of the minimum time
required}\label{estimates-of-the-minimum-time-required}

Across all the populations, I found an average minimum time series
length required (\(T_{min}\)) of 15.9 (SD=8.3), with a wide distribution
(Fig. \ref{fig:min_time_dist}b). Estimates of \(T_{min}\) varied between
biological class (Fig. \ref{fig:min_time_dist}a). Ray-finned fish (class
Actinopterygii) typically had estimates of \(T_{min}\) over 20 years.
Birds (class Aves) had a much wider distribution of \(T_{min}\), but
usually required less years of sampling. Differences between these
classes can be explained by differences in variability in population
size and strength of trends in abundance (Fig. \ref{fig:class}).

This time-frame is in line with past work on a short-lived snail species
(Rueda-Cediel \emph{et al.} 2015) and a long-lived whale species (Gerber
\emph{et al.} 1999). Hatch (2003) used seabird monitoring data to
estimate minimum sampling requirements. He found that the time required
ranged from 11 to 69 years depending on species, trend strength, and
study design.

@@

The minimum time series length required was strongly correlated with
trend strength (i.e.~estimated slope coefficient from linear
regression), coefficient of variation in population size, and
autocorrelation in population size (Figs. \ref{fig:correlates}a-c). All
three of these explanatory variables were significant and had large
effect sizes (see Table A1). Combined, trend strength, coefficient of
variation in population, and autocorrelation account for 75.1\% of the
explained deviance (Zuur \emph{et al.} 2009) in minimum time series
length required.

There was life-history information available (Myhrvold \emph{et al.}
2015) for 547 populations representing 315 different species, all of
which were birds (Aves class). Some life-history traits were significant
predictors for the minimum time required (Figs. \ref{fig:correlates}d-h,
Tables A2,A3). However, even combined, all five of the life-history
traits accounted for only 5.99\% of the explained deviance in minimum
time series length required. In addition, when accounting for trend
strength, coefficient of variation, and autocorrelation, no life-history
traits were significant predictors of the minimum time required (Table
A3).

I initially hypothesized that species with longer lifespans or
generation times may require a longer sampling period. This result could
have been a result of at least two factors. First, the data I used may
not include a diverse enough set of species with different life-history
traits. Second, the question I posed, whether a population is increasing
or decreasing, was specifically concerned with population trends over
time. Therefore, life-history characteristics may be more important for
other questions more closely tied to species biology. For example,
Blanchard \emph{et al.} (2007) used detailed simulations of
spatially-distributed fisheries to compare survey designs. They found
that statistical power depended on survey design, temperature
preferences, and the degree of population patchiness.

@@

For the populations I examined, the IUCN criteria may be overly
simplistic (Fig. \ref{fig:IUCN_analysis}). For many populations, the
IUCN criteria suggest more years than necessary are required to assess a
population as vulnerable. Conversely, for other populations the IUCN
criteria suggest sampling times that are less than the minimum time
required for statistical power. This suggests that the IUCN criteria are
probably too simplistic as the minimum time required does not correlate
with generation time or other biological covariates (Fig.
\ref{fig:correlates}d-h).

\section{Related questions}\label{related-questions}

An important related idea is the optimal allocation of sampling effort
in space versus time. In a theoretical investigation of this question,
Rhodes \& Jonzen (2011) found that the optimal allocation of sampling
depended strongly on temporal and spatial autocorrelation. If spatial
population dynamics were highly correlated, then it was better to sample
more temporally, and vice versa. The empirical data supports this idea
as populations with strong temporal autocorrelation needed less years of
sampling (Fig. \ref{fig:correlates}). Morrison \& Hik (2008) also
studied the optimal allocation of sampling effort in space versus time,
but used emprical data from a long-term survey of the collared pika
(\emph{Ochotona collaris}) in the Yukon. They found that surveys less
than 5 years may be misleading and that extrapolating from one
population to another, even when nearby geographically, may be
untenable.

Seavy \& Reynolds (2007) asked whether statistical power was even a
useful framework for assessing long-term population trends. They used 24
years of census data on Red-tailed Tropicbirds
(\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50\%
decline over 10 years almost always resulted in high statistical power
(above 0.8). Therefore, they cautioned against only using power analyses
to design monitoring schemes and instead argued for metrics that would
increase precision: improving randomization, reducing bias, and
increasing detection probability. Power analyses should not be the only
consideration when designing monitoring schemes. However, unlike Seavy
\& Reynolds (2007), the results here indicate that longer than 10 years
is often needed to achieve high statistical power.

\section{Limitations}\label{limitations}

This paper has some limitations in determining the minimum time series
length required. First, the minimum time estimated is particular to the
specific question of interest. An additional complication is that for
the empirical approach, the subsampling of the full time series allows
for estimates of power, but the individual subsamples are clearly not
independent of one another. In an ideal setting, a specific population
model would be parameterized for each population of interest. Then,
model simulations could be used to estimate the minimum time series
required to address each specific question of interest. Clearly, this is
not always practical, especially if conducting analyses for a wide array
of species.

\section{Conclusions}\label{conclusions}

Power analyses are not a novel aspect of ecological research (Legg \&
Nagy 2006). However, power analyses are still underutilized, especially
in the context of time series analyses. I used a database of 822
populations to determine the minimum time series length required to
detect population trends. This goes beyond previous work that either
focused on theoretical investigations or a limited number of species. I
show that to identify long-term changes in abundance, on average 15.91
years of continuous monitoring are often required (Fig.
\ref{fig:min_time_dist}). However, there is wide distribution of
estimated minimum times. Therefore, it is probably not wise to use a
simple threshold number of years in monitoring design. Further, contrary
to my initial hypotheses, minimum time required did not correlate with
generation time or any other life-history traits (Fig.
\ref{fig:correlates}d-h). This result argues against overly simplified
measures of minimum sampling time based on generation length or other
life-history traits, like those of the IUCN criteria (Fig.
\ref{fig:IUCN_analysis}).

The design of monitoring programs should include calculations of
statistical power, the allocation of sampling in space versus time
(Larsen \emph{et al.} 2001, Rhodes \& Jonzen 2011), and metrics to
increase precision (Seavy \& Reynolds 2007). Ideally, a formal decision
analysis to evaluate these different factors would be conducted to
design or assess any monitoring program (Hauser \emph{et al.} 2006,
McDonald-Madden \emph{et al.} 2010). This type of formal decision
analysis would also include information on the costs of monitoring.
These costs include the actual costs of sampling (Brashares \& Sam 2005)
and the ecological costs of inaction (Thompson \emph{et al.} 2000).

For many populations, short time series are probably not reliable for
detecting population trends. This result highlights the importance of
long-term monitoring programs. From both a scientific and management
perspective, estimates of \(T_{min}\) are important. If a time series is
too short, we lack the statistical power to reliably detect long-term
population trends. In addition, a time series that is too long may be a
poor use of already limited funds (Gerber \emph{et al.} 1999). Further,
more data is not always best in situations where management actions need
to be taken (Martin \emph{et al.} 2012, 2017). When a population trend
is detected, it may be too late for management action. In these
situations, the precautionary principle may be more appropriate
(Thompson \emph{et al.} 2000). Future work should examine other species,
with a wider range of life-history characteristics. In addition, similar
approaches can be used to determine the minimum time series length
required to address additional questions of interest.
